---
title: "rmapshaper Basics"
author: "Andy Teucher"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rmapshaper Basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

rmapshaper is a package which is an R wrapper around the awesome [mapshaper](https://github.com/mbloch/mapshaper/) tool by Matthew Bloch, which has both a [Node.js command-line tool](https://github.com/mbloch/mapshaper/wiki/Introduction-to-the-Command-Line-Tool) as well as an [interactive web tool](http://mapshaper.org/).

The main advantage of the package is the availability of the topologically-aware simplification algorithm in `ms_simplify` (provided by the `simplify` tool in mapshaper). This means that shared boundaries between adjacent polygons are always kept intact, with no gaps or overlaps, even at high levels of simplification. It uses the  [Visvalingam](http://bost.ocks.org/mike/simplify/) simplification method.

At this time, `rmapshaper` provides the following functions:

- `ms_simplify` - simplify polygons or lines
- `ms_clip` - clip an area out of a layer using a polygon layer or a bounding box. Works on polygons, lines, and points
- `ms_erase` - erase an area from a layer using a polygon layer or a bounding box. Works on polygons, lines, and points
- `ms_dissolve` - aggregate polygon features, optionally specifying a field to aggregate on. If no field is specified, will merge all polygons into one.
- `ms_explode` - convert multipart shapes to single part. Works with polygons, lines, and points in geojson format, but currently only with polygons and lines in the `Spatial` classes (not `SpatialMultiPoints` and `SpatialMultiPointsDataFrame`).
- `ms_lines` - convert polygons to topological boundaries (lines)
- `ms_innerlines` - convert polygons to shared inner boundaries (lines)
- `ms_points` - create points from a polygon layer
- `ms_filter_fields` - Remove fields from the attributes
- `ms_filter_islands` - Remove small detached polygons

This short vignette focuses on simplifying polygons with the `ms_simplify` function.

### Usage

rmapshaper works with geojson strings (character objects of class `geo_json`) and `list` 
geojson objects of class `geo_list`. These classes are defined in the `geojsonio` 
package. It also works with `Spatial` classes from the `sp` package.

We will use the `states` dataset from the `geojsonio` package and first turn it 
into a `geo_json` object:

```{r}
library(geojsonio)
library(rmapshaper)
library(sp)

states_json <- geojson_json(states, geometry = "polygon", group = "group")
```

For ease of illustration via plotting, we will convert to a `SpatialPolygonsDataFrame`:
```{r}
states_sp <- geojson_sp(states_json)

## Plot the original
plot(states_sp)
```

Now simplify using default parameters, then plot the simplified states
```{r}
states_simp <- ms_simplify(states_sp)
plot(states_simp)
```

You can see that even at very high levels of simplification, the mapshaper
simplification algorithm preserves the topology, including shared boundaries:

```{r}
states_very_simp <- ms_simplify(states_sp, keep = 0.001)
plot(states_very_simp)
```

Compare this to the output using `rgeos::gSimplify`, where overlaps and gaps are evident:

```{r}
library(rgeos)
states_gsimp <- gSimplify(states_sp, tol = 1, topologyPreserve = TRUE)
plot(states_gsimp)
```

The package also works with `sf` objects. This time we'll demonstrate the `ms_innerlines` function:

```{r}
library(sf)
states_sf <- st_as_sf(states_sp)
states_sf_innerlines <- ms_innerlines(states_sf)
plot(states_sf_innerlines)
```

All of the functions are quite fast with `geo_json` character objects and `geo_list`
list objects. They are slower with the `Spatial` classes due to internal conversion 
to/from json. If you are going to do multiple operations on large `Spatial` objects, 
it's recommended to first convert to json using `geojson_list` or `geojson_json` from 
the `geojsonio` package. All of the functions have the input object as the first argument, 
and return the same class of object as the input. As such, they can be chained together. 
For a totally contrived example, using `states_sp` as created above:

```{r eval=rmapshaper:::check_v8_major_version() >= 6}
library(geojsonio)
library(rmapshaper)
library(sp)
library(magrittr)

## First convert 'states' dataframe from geojsonio pkg to json
states_json <- geojson_json(states, lat = "lat", lon = "long", group = "group", 
                            geometry = "polygon")

states_json %>% 
  ms_erase(bbox = c(-107, 36, -101, 42)) %>% # Cut a big hole in the middle
  ms_dissolve() %>% # Dissolve state borders
  ms_simplify(keep_shapes = TRUE, explode = TRUE) %>% # Simplify polygon
  geojson_sp() %>% # Convert to SpatialPolygonsDataFrame
  plot(col = "blue") # plot
```

### Using the system mapshaper

Sometimes if you are dealing with a very large spatial object in R, `rmapshaper`
functions will take a very long time or not work at all. As of version `0.4.0`, 
you can make use of the system `mapshaper` library if you have it installed.
This will allow you to work with very large spatial objects.

First make sure you have mapshaper installed:

```{r eval=nzchar(Sys.which("mapshaper"))}
check_sys_mapshaper()
```

If you get an error, you will need to install mapshaper. First install node 
(https://nodejs.org/en/) and then install mapshaper with:

```
npm install -g mapshaper
```

Then you can use the `sys` argument in any rmapshaper function:

```{r eval=nzchar(Sys.which("mapshaper"))}
states_simp_internal <- ms_simplify(states_sf)
states_simp_sys <- ms_simplify(states_sf, sys = TRUE)

all.equal(states_simp_internal, states_simp_sys)
```
